A flexible integer linear programming formulation for scheduling clinician on-call service in hospitals

### 2019-09-14 paper/paper.problem.tex
editing problem

At St.\ Michael's Hospital, the division of infectious diseases (ID) offers
general ID and HIV consultation on inpatients as two parallel services. Each
service provides clinical care throughout the year, during regular work weeks as
well as weekends and holidays. The clinicians in the division typically receive
a schedule in advance, outlining their on-call service dates for the full year.
In the yearly schedule each clinician is assigned to blocks of regular work
weeks and weekends. Each block corresponds to two consecutive work weeks. Apart
from long (holiday) weekends, a work week starts on Monday at 8 A.M. and ends on
Friday at 5 P.M. Accordingly, weekend service starts on Friday at 5 P.M., and
ends on Monday at 8 A.M. During the weekend, ID and HIV consultation services
are combined and provided by one clinician. During the regular work week the ID
and HIV services are led by one clinician each.

Several constraints are placed on the on-call assignments. First, each clinician
has limits on the number of blocks they can and must work during the year,
depending on the type of consultation. For instance, a clinician might have to
provide 3-5 blocks of general ID consultation as well as 2-3 blocks of HIV
consultation throughout the year. These limits may change from year to year as
the number of clinicians in the department changes. Moreover, the schedule
should not assign a clinician to work for two consecutive blocks or two
consecutive weekends in a row. The schedule should also distribute both regular
and holiday weekends equally among all clinicians.

In addition to maintaining a balanced work load among clinicians, the schedule
should also accommodate their preferences. Clinicians provide their requests for
time off ahead of schedule generation so that the requests may be integrated
into the schedule. Individuals may specify days, weeks or weekends off, with the
understanding that any blocks overlapping with their request will be assigned to
a different clinician where possible. For example, if a clinician only requests
a given Monday and Tuesday off, the schedule will generally avoid assigning the
entire block to that clinician. Clinicians also prefer to have their weekend and
block assignments close together, so the schedule should account for this when
distributing assignments. A summary of the outlined constraints is given in
Table~\ref{tbl:constraint-summary}.

@@

In most scheduling problems, the constraints can be divided into hard and soft
constraints. Hard constraints must be satisfied by any candidate solution, while
soft constraints can be used to select a more favourable solution from all
candidate solutions. Typically, soft constraints are encoded as objective
functions rather than actual constraints. These objective functions are
maximized or minimized when finding a solution. In the case of our clinician
scheduling problem, we chose Block Requests, Weekend Requests and Block-Weekend
Adjacency as our soft constraints, while the rest of the constraints are hard.
Though it is important to take clinician requests into account when constructing
the schedule, it is crucial that the work-load of the schedule is balanced among
all clinicians, and the needs of the patients are fulfilled. It is important to
note that - irrespective of how the schedule is generated - clinicians may
exchange certain weeks or days throughout the year after the schedule is
implemented. The approach to solving the clinician scheduling problem does not
account for these future exchanges, and only focuses on the full time horizon.

>>>>

At St.\ Michael's Hospital, the division of infectious diseases (ID) offers
separate but concurrent services for general ID and HIV consultation. Each
service provides clinical care throughout the year, during regular working hours
and on weekends and holidays. The schedule is created in advance,
outlining all work-week and weekend shifts for the full year.
In the yearly schedule clinicians are assigned to ``blocks'' of
two consecutive regular work weeks and individual weekends. Apart
% JK: when I first read this I thought weekends were also in 2-week blocks
from long (holiday) weekends, a work week starts on Monday at 8 A.M. and ends on
Friday at 5 P.M. Accordingly, weekend service starts on Friday at 5 P.M., and
ends on Monday at 8 A.M. During the weekend, ID and HIV consultation services
are combined and provided by one clinician. During the regular work week the ID
and HIV services are led by one clinician each.

Two types of constraints are placed on the clinician assignments. 
% JK: I think you need to introduce BC and WC constraints here,
%     since they appear in Table 1 without text in the body.
First, each clinician
must work between a minimum and maximum number of blocks of each service during the year.
For instance, one clinician might have to
provide 3-5 blocks of general ID service and 2-3 blocks of HIV service
% JK: trying to reduce synonyms "service" / "consultation"
throughout the year. These limits may be different fo each clinician,
and they may change from year to year as
the number of clinicians in the department changes. Second, the schedule
should not assign a clinician to work for two consecutive blocks or two
consecutive weekends. The schedule should also distribute regular
weekends and holiday weekends each equally among all clinicians.

In addition to balancing the work load among clinicians, the schedule
should accommodate their preferences. Clinicians provide their requests for
time off before schedule generation so that the requests may be integrated
into the schedule. Clinician may specify days, weeks or weekends off, with the
understanding that any blocks overlapping with their request will be assigned to
a different clinician where possible. For example, if a clinician only requests
a given Monday and Tuesday off, the schedule should avoid assigning the
entire block to that clinician. Clinicians also typically prefer to have their weekend and
block assignments side by side, so the schedule should accommodate this where possible.
A summary of the outlined constraints is given in
Table~\ref{tbl:constraint-summary}.

@@

In most scheduling problems, the constraints can be divided into hard and soft
constraints. Hard constraints must be satisfied by any candidate solution, while
soft constraints can be used to select a more favourable solution from a set of
candidate solutions. Typically, soft constraints are therefore encoded as objective
functions which are
maximized when finding a solution. In the case of the clinician
% JK: the direction of max / minimization is arbitrary,
%     I think sufficient to pick one here to avoid confusion.
scheduling problem described above, we chose Block Requests, Weekend Requests and Block-Weekend
Adjacency as our soft constraints, while the rest of the constraints are hard.
While it is important to take clinician requests into account when constructing
the schedule, it is crucial that the work-load of the schedule is balanced among
all clinicians, and all services are assigned.
After the schedule is generated, clinicians may
exchange certain weeks or days throughout the year.
Our approach to schedule generation does not
consider such future exchanges, and instead generates on one full year at a time.
% JK: I think this para (4) should go instead as the 2nd in this section.
%     you could introduce the ideas of hard and soft constraints (this para) first,
%     and then have the current paras (2) and (3) after this one, where the reader
%     is already aware that (2) is hard constraints and (3) is soft.
%     Oh, I see that they're not currently split up exactly, but I think it would be
%     helpful to do so. Maybe double check with SM though.
